---
title: "Course Project 1"
author: "Jim Milks"
date: "1/8/2021"
output:
  md_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

I loaded the data then transformed the date variable from character to date class

```{r}
library(stats)
activity <- read.csv("activity.csv", header = T)
activity$date <- strptime(activity$date, "%Y-%m-%d")
summary(activity)
```

## What is mean total number of steps taken per day?
Using the lubridate and dplyr packages, I first summed the steps per day, then created a histogram of the total steps per day. I then found the mean number of steps taken per day and the median number of steps taken per day.

```{r}
library(lubridate)
library(dplyr)

#Sum the total steps per day
total_steps_per_day <- activity %>% group_by(day = floor_date(date, "day")) %>% summarize(steps = sum(steps, na.rm = T))

#Create a histogram of total steps per day
hist(total_steps_per_day$steps, xlab = "Total steps per day", main = "Histogram of total steps taken per day")

#Find the mean and median
mean(total_steps_per_day$steps)
median(total_steps_per_day$steps)
```
The mean was 9,354.23 steps per day whereas the median was higher at 10,395 steps per day.

## What is the average daily activity pattern?
After looking at each individual day, I then examined the steps per 5 minute time interval. I calculated the average number of steps for each interval across all days, created a time-series plot showing the activity at each interval, then located the interval with the greatest average activity level. Interval 835 with an average of 206 steps had the greatest average number of steps.

```{r}
library(dplyr)

#Find the mean steps per time interval
mean_steps_per_interval <- activity %>% group_by(interval) %>% summarize(steps = mean(steps, na.rm = T))

#Create a time series plot of average steps vs time
plot(steps~interval, data = mean_steps_per_interval, type = "l", xlab = "Time interval (minutes)", ylab = "Mean number of steps", main = "Mean number of steps vs Time", lwd = 1.5)

#Locate the time interval with the highest mean steps
which.max(mean_steps_per_interval$steps)
mean_steps_per_interval[104,]
```
Interval 835 with an average of 206 steps had the greatest average number of steps.

## Imputing missing values
Given the number of missing values, there's a chance there may be some bias in the calculations due to missing values. I filled in the missing values with the average steps per time interval, then re-ran the analysis for daily steps, mean steps per day, and median steps per day.

```{r}
library(dplyr)
library(lubridate)
library(zoo)

#Find the number of NAs in the data
sum(is.na(activity$steps))

#Clone the dataset, fill in the missing values in the new data frame, and check to ensure all the NA have been replaced
activity_filled <- activity
activity_filled$steps <- ave(activity_filled$steps, activity_filled$interval, FUN = na.aggregate)
sum(is.na(activity_filled$steps))

#Create a histogram of the filled-in data
total_steps_per_day_imputed <- activity_filled %>% group_by(day = floor_date(date, "day")) %>% summarize(steps = sum(steps, na.rm = T))
hist(total_steps_per_day_imputed$steps, xlab = "Total steps per day", main = "Histogram of total steps per day with imputed data")

#Find the mean and median of the imputed data set
mean(total_steps_per_day_imputed$steps, na.rm = T)
median(total_steps_per_day_imputed$steps, na.rm = T)
```
Imputing data into the data frame increased the mean from 9,354.23 to 10,766.16 and the median from 10,395 to 10,766.19. The overall effect was to normalize the total steps per day data, correcting the skew in the original data.

## Are there differences in activity patterns between weekdays and weekends?
I then examined whether there were differences in activity patterns between weekdays and weekends using the filled-in data set. I started by first splitting the dataset into weekdays and weekends before plotting the average steps per time interval.
```{r}
library(dplyr)
library(ggplot2)

#Create new variable to pull weekday names and then subset the dataset based on weekday or weekend
activity_filled$weekday <- weekdays(activity_filled$date, abbreviate = FALSE)
weekday.1 <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity_filled$week <- factor((activity_filled$weekday %in% weekday.1), 
                               levels = c(FALSE, TRUE), labels = c("weekend", "weekday"))

#Calculate mean steps per time interval for weekdays and weekends
mean_steps_interval_weekdays <- activity_filled %>% group_by(week, interval) %>% summarize(steps = mean(steps))

mean_steps_interval_weekdays %>% group_by(week) %>% summarize(steps = mean(steps))
mean_steps_interval_weekdays %>% group_by(week) %>% summarize(steps = median(steps))
mean_steps_interval_weekdays %>% group_by(week) %>% summarize(steps = max(steps))

#Plot mean steps per time interval split by weekday vs weekend
ggplot(data = mean_steps_interval_weekdays, aes(x = interval, y = steps)) +
        geom_line() +
        theme_bw() +
        facet_grid(week~.) +
        labs(title = "Mean steps per time interval",
             subtitle = "Weekends vs Weekdays",
             xlab = "Time (minutes)",
             ylab = "Mean steps per interval")
```

Although the weekday maximum is higher (230 steps vs 167), the weekend had the higher average number of steps (42.4 vs 35.6) and higher median (32.3 vs 25.8). While at first glance at the time series plot appears to show that weekdays had more activity, it's actually the weekends that showed the higher activity levels, thereby demonstrating that looks can be deceiving, even in data analysis.